\protect Connecting every bit of knowledge: \\ The Structure of Wikipedia's first link network

### 2016-02-07 writeup/wikipedia-network.tex
first pass syntax edits complete

TODO: 
* fix end of second paragraph in Results Section
    * power-law biz and total peter is asking for
* fix exponent reporting (with alpha) 
    * figure out what gamma size freq means

@@

The authors of a Wikipedia article choose where and whether to include a 
reference to another Wikipedia article
in the HTML markup.
For example, the authors of the 
\href{https://en.wikipedia.org/wiki/Train}{"Train"}
article chose
"Rail Transport," "Cargo," and "Passengers" 
as pertinent articles to reference in describing trains
\cite{wiki_train}
.
Here we focus on the first hyperlink to another Wikipedia article, which 
marks the earliest moment in the body text where the authors
chose to connect the information discussed to that of another article.
By following the first hyperlink to another article in the English edition of
Wikipedia as of November, 2014, we connect the ideas in one article to those in another, ultimately forming a directed network: 
{\it Wikipedia's First Link Network} (FLN).

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\textwidth]{graphics/Train.pdf}  
  \caption{
    \textbf{First Link Path For "Train".}
    We follow the First Link to another Wikipedia article
    in the main body of the article---the area
    inside the green rectangle, which excludes 
    side bar elements, the navigation bar and title. 
    In this example, the First Link to another Wikipedia article is "Rail Transport." 
    We can again select the First Link on the "Rail Transport"
    article, repeating the process to 
    form a path of First Links.
  }
  \label{fig:Train First Links}
\end{figure*}

What information could the First Links possibly reveal?
Inspired by the claim that the majority of First Links lead to the 
"Philosophy" article---popularized by an 
\href{https://xkcd.com/903/}{xkcd}
comic and subsequently
discussed in blog posts 
\cite{mat_blog, Ilmari_first_links, xkcd}
---we holistically study 
Wikipedia's FLN as a map relating areas of human knowledge. 
The map is a hierarchical taxonomy where, for example, "Train" links to a parent node, 
"Rail Transport," while many child nodes, "Steel" and "Horsepower," link to 
"Train." Unlike previous taxonomies
created by individuals
\cite{locke, descartes, aristotle}
or a select group of individuals 
\cite{hist_thesaurus}, 
the organization of ideas in the FLN 
does not emerge from a centralized effort. 
Rather each link in the FLN emerges from the choice of each author and ensuing edits.
Collectively, the FLN is a wealth of relations among inventions, places,
figures, objects, and events across space and time.

@@

The "Philosophy" article despite outranking every article by traversal visits,
has only $240$ thousand page views in October compared to the $1.4$ million
page views for "United States".
While "Philosophy" is not itself the most popular article, "Philosophy" serves as 
an accumulation point for ideas. Wikipedia users do not login specifically
to read about "Philosophy," but the authors relate ideas to "Philosophy" 
by ultimately anchoring the flow of First Links to "Philosophy."

@@

\section{Reflections}

The findings here should only be considered within the limitations of their context.
We examined only the English version of Wikipedia at a particular moment in time.
Furthermore, we only studied the First Link in the main body  of each article
as a means to related one article to another. Finally, Wikipedia, while the largest 
collection of human knowledge, is rife with the biases of the many contributing editors
\cite{bias}
. Nevertheless, the findings do reveal
generalizable relationships, point to foundational notions, and uncover many curiosities.

Among the curiosities is the multiple appearance of scale-free distributions within the network. 
The three metrics we developed: path length, traversal visits, and traversal funnels are all marked 
by scale-free distributions. Few articles have most traversal visits, few paths have an exceptionally long path length, and even fewer
articles are responsible for funneling most paths. When measured against the traversal funnels, 
"Philosophy" emerges as an exceptional article by orders of magnitude. 
Nevertheless, many other foundational ideas emerged naturally within the First Link Network. 
Basins around "Community", "State", and "Science" reveal a foundational structure within the network. 
More curious is the emergence of recently prominent political and economics topics such as "Fossil Fuel" and "Health Care" 
within the highest ranking funnels. 
Wikipedia seems to reflect not only timeless foundations, but also the topical (at least within English speaking society).

Future work could examine other language versions of Wikipedia for potentially telling cultural or regional differences as well as expand the network to more than the First Link.
These findings also form the basis for the creation of a taxonomy where 
every idea, event, or object sits within a hierarchy of connected notions.
The taxonomy would extend a traditional word thesaurus beyond mere synonyms to a related hierarchy of concepts.
Applications could range from an enhanced thesaurus of ideas to psychological insights into how humans form associations.
Specifically, an ever-evolving reference of related hierarchical concepts can be applied to search engine algorithms 
or natural language processing.

\acknowledgments
Thank you to my friend RJ for pointing out the Reddit post 
\cite{reddit}
describing how the majority of links lead to "Philosophy," inspiring this research.

>>>>

TODO: 
\begin{itemize}
    \item fix end of second paragraph in Results Section\\
    * power-law biz and total peter is asking for
\item fix exponent reporting (with alpha) \\
    * figure out what gamma size freq means
\end{itemize}

@@

The authors of a Wikipedia article choose where and whether to include a 
reference to another Wikipedia article
in the HTML markup.
For example, the authors of the 
\href{https://en.wikipedia.org/wiki/Train}{``Train''}
article chose
``Rail Transport,'' ``Cargo,'' and ``Passengers'' 
as pertinent articles to reference in describing trains
\cite{wiki_train}
.
Here we focus on the first hyperlink to another Wikipedia article, which 
marks the earliest moment in the body text where the authors
chose to connect the information discussed to that of another article.
By following the first hyperlink to another article in the English edition of
Wikipedia as of November, 2014, we connect the ideas in one article to those in another, ultimately forming a directed network: 
{\it Wikipedia's First Link Network} (FLN).

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\textwidth]{graphics/Train.pdf}  
  \caption{
    \textbf{First Link Path For ``Train.''}
    We follow the First Link to another Wikipedia article
    in the main body of the article---the area
    inside the green rectangle, which excludes 
    side bar elements, the navigation bar and title. 
    In this example, the First Link to another Wikipedia article is ``Rail Transport.'' 
    We can again select the First Link on the ``Rail Transport''
    article, repeating the process to 
    form a path of First Links.
  }
  \label{fig:Train First Links}
\end{figure*}

What information could the First Links possibly reveal?
Inspired by the claim that the majority of First Links lead to the 
``Philosophy'' article---popularized by an 
\href{https://xkcd.com/903/}{xkcd}
comic and subsequently
discussed in blog posts 
\cite{mat_blog, Ilmari_first_links, xkcd}
---we holistically study 
Wikipedia's FLN as a map relating areas of human knowledge. 
The map is a hierarchical taxonomy where, for example, ``Train'' links to a parent node, 
"Rail Transport," while many child nodes, "Steel" and "Horsepower," link to 
"Train." Unlike previous taxonomies
created by individuals
\cite{locke, descartes, aristotle}
or a select group of individuals 
\cite{hist_thesaurus}, 
the organization of ideas in the FLN 
does not emerge from a centralized effort. 
Rather each link in the FLN emerges from the choice of each author and ensuing edits.
Collectively, the FLN is a wealth of relations among inventions, places,
figures, objects, and events across space and time.

@@

The "Philosophy" article, despite outranking every article by traversal visits,
has only $240$ thousand page views in October compared to the $1.4$ million
page views for "United States".

@@

\section{Concluding Remarks}

The findings here should only be considered within the limitations of their context.
We examined only the English editiion of Wikipedia at a particular moment in time, 
considering only the first link in in each article as a means to related one article to another.
Finally, Wikipedia, while the largest 
collection of human knowledge, is rife with the biases of the many contributing editors
\cite{bias}
. Nevertheless, the findings do reveal
generalizable relationships and point to foundational notions.

Among our observations is the multiple appearance of scale-free distributions within the network. 
Few articles have most traversal visits, few paths have an exceptionally long path length, and even fewer
articles are responsible for funneling most paths. When measured against the traversal funnels, 
"Philosophy" emerges as an exceptional article by orders of magnitude. 
Nevertheless, many other foundational ideas emerged naturally within the First Link Network. 
Basins around "Community", "State", and "Science" reveal a foundational structure within the network. 
More curious is the emergence of recently prominent political and economics topics such as "Fossil Fuel" and "Health Care" 
within the highest ranking funnels. 
Wikipedia seems to reflect not only timeless foundations, but also the topical (at least within English speaking society).

Future work could examine other language versions of Wikipedia for potentially telling cultural or regional differences as well as expand the network to more than the First Link.
These findings also form the basis for the creation of a taxonomy where 
every idea, event, or object sits within a hierarchy of connected notions.
The taxonomy would extend a traditional word thesaurus beyond mere synonyms to a related hierarchy of concepts.
Applications could range from an enhanced network of ideas to psychological insights into how humans form associations.
Specifically, an ever-evolving reference of related hierarchical concepts could be applied to search engine algorithms 
or natural language processing.

\acknowledgments
MI is grateful to RJ for pointing out the Reddit post 
\cite{reddit}
describing how the majority of links lead to "Philosophy," inspiring this research.
